Viral Sequence Sorting Tools Evaluation

### 2022-10-30 EnvironmentalData/DrinkingWaterMetagenomes/identifying_viral_contigs_environmental_dw.Rmd
New testing set visualizations.

```{r}
ggplot(num_viruses, aes(x=toolcombo, y=num_viruses, 
                                  color=numtools, fill=numtools)) +
  geom_point() +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tv, DVF, ntv, VB, VS, VS2)") +
  ylab("Num Viruses Predicted")
```

@@

```{r}
ggplot(num_viruses, aes(x=toolcombo, y=prop_vir, 
                                  shape=numtools, color=assembly, fill=assembly)) +
  geom_point() +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tv, DVF, ntv, VB, VS, VS2)") +
  ylab("Proportion of Contigs >3kb Viral")
```

@@

```{r}
num_viruses_uknl <- num_viruses[num_viruses$assembly %in% c("Aberdeen_Man",
                                                            "Amsterdam_Leid",
                                                            "Amsterdam_Wee",
                                                            "Birmingham_Maria",
                                                            "Cardiff_Maria",
                                                            "Dundee_Clat",
                                                            "Dundee_Maria",
                                                            "Edinburgh_Glen",
                                                            "Exeter_Maria",
                                                            "Glasgow_Bal",
                                                            "Glasgow_Mil",
                                                            "London_Maria",
                                                            "Nieuwkoop_DHB",
                                                            "South_Holland_Dla",
                                                            "Spannenburg"),]
```

```{r}
ggplot(num_viruses_uknl, aes(x=toolcombo, y=prop_vir, 
                                  shape=numtools, color=assembly, fill=assembly)) +
  geom_point() +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tv, DVF, ntv, VB, VS, VS2)") +
  ylab("Proportion of Contigs >3kb Viral")
```

>>>>

```{r}
ggplot(num_viruses, aes(x=toolcombo, y=num_viruses, 
                                  color=numtools, fill=numtools)) +
  geom_point() +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tnv, DVF, tv, VB, VS, VS2)") +
  ylab("Num Viruses Predicted")
```

@@

```{r}
ggplot(num_viruses, aes(x=toolcombo, y=prop_vir, 
                                  shape=numtools, color=assembly, fill=assembly)) +
  geom_point() +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tnv, DVF, tv, VB, VS, VS2)") +
  ylab("Proportion of Contigs >3kb Viral")
```

only plotting the average

```{r}
num_viruses$numtools <- as.numeric(num_viruses$numtools)

median_prop_vir_by_tool <- num_viruses %>% 
  select(toolcombo, prop_vir, numtools) %>%
  group_by(toolcombo) %>%
  summarize(median=median(prop_vir),
         #sd=sd(prop_vir),
         numtools=mean(numtools))

median_prop_vir_by_tool$toolcombo <- factor(median_prop_vir_by_tool$toolcombo, 
                                          levels = median_prop_vir_by_tool$toolcombo[order(median_prop_vir_by_tool$median, decreasing = F)])

median_prop_vir_by_tool$numtools <- as.factor(median_prop_vir_by_tool$numtools)
```

```{r}
ggplot(median_prop_vir_by_tool, aes(y=median, x=toolcombo,
                                  color=numtools, fill=numtools, shape=numtools))  +
  geom_hline(yintercept=0, color="grey") +
  geom_point() +
#  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2,
#                 position=position_dodge(.9)) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tnv, DVF, tv, VB, VS, VS2)") +
  ylab("Proportion of Contigs >3kb Viral")
```
```{r}
ggplot(median_prop_vir_by_tool, aes(x=numtools, y=median, 
                                  color=numtools, fill=numtools)) +
  geom_boxplot() +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  ylab("Proportion Viruses (%)") +
  xlab("Number of Tools") +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1))
```

@@

```{r}
num_viruses_uknl <- num_viruses[num_viruses$assembly %in% c("Aberdeen_Man",
                                                            "Amsterdam_Leid",
                                                            "Amsterdam_Wee",
                                                            "Birmingham_Maria",
                                                            "Cardiff_Maria",
                                                            "Dundee_Clat",
                                                            "Dundee_Maria",
                                                            "Edinburgh_Glen",
                                                            "Exeter_Maria",
                                                            "Glasgow_Bal",
                                                            "Glasgow_Mil",
                                                            "London_Maria",
                                                            "Nieuwkoop_DHB",
                                                            "South_Holland_Dla",
                                                            "Spannenburg"),]

num_viruses_uknl$residual <- "residual"
num_viruses_uknl$residual[num_viruses_uknl$assembly=="Amsterdam_Leid" |
                          num_viruses_uknl$assembly=="Amsterdam_Wee" |
                          num_viruses_uknl$assembly=="Nieuwkoop_DHB" |
                          num_viruses_uknl$assembly=="South_Holland_Dla" |
                          num_viruses_uknl$assembly=="Spannenburg"
                          ] <- "no residual"
```

```{r}
num_viruses_uknl$numtools <- as.numeric(num_viruses_uknl$numtools)

mean_prop_vir_by_tool <- num_viruses_uknl %>% 
  select(toolcombo, prop_vir, numtools, residual) %>%
  group_by(toolcombo, residual) %>%
  summarize(prop_vir=mean(prop_vir),
         #sd_per_vir=sd(prop_vir),
         numtools=mean(numtools))

mean_uknl <- mean_prop_vir_by_tool %>% 
  group_by(toolcombo) %>%
  summarize(mean=mean(prop_vir))

mean_prop_vir_by_tool$toolcombo <- factor(mean_prop_vir_by_tool$toolcombo, 
                                          levels = mean_uknl$toolcombo[order(mean_uknl$mean, decreasing = F)])

mean_prop_vir_by_tool$numtools <- as.factor(mean_prop_vir_by_tool$numtools)
```

```{r}
num_viruses_uknl$toolcombo <- factor(num_viruses_uknl$toolcombo, 
                                          levels = levels(mean_prop_vir_by_tool$toolcombo))
num_viruses_uknl$numtools <- as.factor(num_viruses_uknl$numtools)

```{r}
ggplot(num_viruses_uknl) +
  geom_hline(yintercept=0, color="grey") +
  geom_point(aes(color=numtools, fill=numtools, y=prop_vir, x=toolcombo, shape=residual), alpha=0.5) +
  geom_point(data=mean_prop_vir_by_tool, aes(x=toolcombo, 
                                               y=prop_vir,
                                             shape=residual),
             color="black") +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tnv, DVF, tv, VB, VS, VS2)") +
  ylab("Proportion of Contigs >3kb Viral") +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1))
```
```{r}
t.test(num_viruses_uknl$prop_vir[num_viruses_uknl$residual=="residual"],
       num_viruses_uknl$prop_vir[num_viruses_uknl$residual=="no residual"]
       )
```

### 2022-10-30 TestingPipeline/Scripts/rule_tuning_2022.09.06.Rmd
New testing set visualizations.

@@

```{r}
viruses$size_class <- "3-10kb"
viruses$size_class[viruses$checkv_length>10000] <- ">10kb"
```

@@

```{r}
ggplot(confusion_by_taxa, aes(x=count, y=as.factor(index),
                   fill=confusion_matrix,
                   color=confusion_matrix)) +
  geom_bar(stat="identity") +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(4)), 0.5),
                    labels=c("false negative", "false positive", 
                             "true negative", "true positive")) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(4)), 1),
                    labels=c("false negative", "false positive", 
                             "true negative", "true positive")) +
  xlab("Number of Sequences") +
  ylab("") + 
  facet_wrap(~seqtype, scales = "free") +
  coord_flip()
```
percent viral

visualizing confusion matrix by taxa
```{r}
confusion_by_taxa <- viruses %>% count(confusion_matrix_high_MCC, seqtype, size_class, Index)

@@

confusion_vir_called <- confusion_vir_called %>% group_by(seqtype, size) %>%
  summarise(mean=mean(per_viral), 
            sd=sd(per_viral))
```

```{r}
ggplot(confusion_vir_called, aes(y=mean, x=size,
                   fill=seqtype,
                   color=seqtype)) +
  geom_bar(stat="identity", position=position_dodge()) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2,
                 position=position_dodge(.9)) +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1)) +
  xlab("Length (bp)") +
  ylab("Sequences Called Viral (%)") 
```

@@

>>>>

```{r}
accuracy_scores_melt <- accuracy_scores %>% 
  select(testing_set_index, precision, recall, MCC, numtools, toolcombo) %>%
  pivot_longer(cols=c(precision, recall, MCC), 
               names_to="performance_metric",
               values_to="performance_metric_score")
```

```{r}
ggplot(accuracy_scores_melt, aes(x=numtools, y=performance_metric_score, 
                                  color=numtools, fill=numtools)) +
  geom_boxplot() +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  ylab("MCC") +
  xlab("Number of Tools") +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1)) +
  facet_wrap(~performance_metric)

```
comparing metric with and without tuning rules
```{r}
accuracy_scores_melt$tuning_inc <- "no"
accuracy_scores_melt$tuning_inc[substring(accuracy_scores_melt$toolcombo, 1, 1)==1] <- "tv"
accuracy_scores_melt$tuning_inc[substring(accuracy_scores_melt$toolcombo, 3, 3)==1] <- "tnv"
accuracy_scores_melt$tuning_inc[substring(accuracy_scores_melt$toolcombo, 1, 1)==1 &
                                  substring(accuracy_scores_melt$toolcombo, 3, 3)==1] <- "both"
```

```{r}
ggplot(accuracy_scores_melt, aes(x=tuning_inc, y=performance_metric_score)) +
  geom_boxplot() +
  geom_point(aes(color=numtools, fill=numtools), alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  ylab("MCC") +
  xlab("Number of Tools") +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1)) +
  facet_wrap(~performance_metric)

ggplot(accuracy_scores_melt, aes(x=tuning_inc, y=performance_metric_score,
                                 color=numtools, fill=numtools)) +
  geom_boxplot() +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  ylab("MCC") +
  xlab("Number of Tools") +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1)) +
  facet_wrap(~performance_metric)

ggplot(accuracy_scores_melt, aes(x=tuning_inc, y=performance_metric_score)) +
  geom_boxplot() +
  geom_boxplot(aes(color=numtools, fill=numtools)) +
 # geom_point(aes(color=numtools, fill=numtools), alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  ylab("MCC") +
  xlab("Number of Tools") +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1)) +
  facet_wrap(~performance_metric)

@@

@@

```{r}
ggplot(confusion_by_taxa, aes(x=count, y=as.factor(index),
                   fill=confusion_matrix,
                   color=confusion_matrix)) +
  geom_bar(stat="identity") +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(4)), 0.5),
                    labels=c("false negative", "false positive", 
                             "true negative", "true positive")) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(4)), 1),
                    labels=c("false negative", "false positive", 
                             "true negative", "true positive")) +
  xlab("Number of Sequences") +
  ylab("") + 
  facet_wrap(~seqtype, scales = "free") +
  coord_flip()
```

differences based on genome size
```{r}
viruses$size_class <- "3-5kb"
viruses$size_class[viruses$checkv_length>5000] <- "5-10kb"
viruses$size_class[viruses$checkv_length>10000] <- ">10kb"
```

```{r}
confusion_by_taxa <- viruses %>% count(confusion_matrix_high_MCC, seqtype, size_class, Index)

@@

confusion_vir_called <- confusion_vir_called %>% group_by(seqtype, size) %>%
  summarise(mean=mean(per_viral), 
            sd=sd(per_viral))

confusion_vir_called$size <- factor(confusion_vir_called$size,
                                    levels = c("3-5kb", "5-10kb", ">10kb"))
```

```{r}
ggplot(confusion_vir_called, aes(y=mean, x=size,
                   fill=seqtype,
                   color=seqtype)) +
  geom_bar(stat="identity", position=position_dodge()) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2,
                 position=position_dodge(.9)) +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1)) +
  xlab("Length") +
  ylab("Sequences Called Viral (%)") 
```

@@

## clustering
```{r}
viral_scores <- matrix(data=0, nrow=nrow(viruses), ncol=nrow(combos_list))
num_viruses <- data.frame(toolcombo=rep(0, nrow(combos_list)),
                          num_viruses=rep(0, nrow(combos_list)))

for (i in 1:nrow(combos_list)) {
  viral_scores[,i] <- getting_viral_set_1(viruses, include_vibrant = combos_list$VIBRANT[i],
                                            include_virsorter = combos_list$VS[i],
                                            include_virsorter2 = combos_list$VS2[i],
                                            include_tuning_viral = combos_list$tune_viral[i],
                                            include_tuning_not_viral = combos_list$tune_not_viral[i],
                                            include_deepvirfinder = combos_list$DVF[i])

  if (max(viral_scores[,i])<=0) {
    num_viruses$num_viruses[i] <- 0
  }
  else {
    num_viruses$num_viruses[i] <- table(viral_scores[,i]>=1)[[2]]
  }

colnames(viral_scores_nozeros) <- num_viruses$toolcombo
```

rownames(tooldata) <- tooldata$toolcombo
```

phyloseq::plot_ordination(physeq = physeq_pooled, ordination = ordination,
                          shape="numtools", color="num_viruses") + 
  geom_point(size = 3) +
  theme_bw()
```

myclusters <- cutree(clusters, h=0.8)
```

```{r}
#names(myclusters[myclusters==1])
#names(myclusters[myclusters==2])
#names(myclusters[myclusters==3])
#names(myclusters[myclusters==4])
#names(myclusters[myclusters==5])

myclusters_df <- separate(myclusters_df, col=combo, into=c("tnv", "DVF",
                                                            "tv", "VB",
                                                            "VS", "VS2"),
                          sep=" ", remove = F)

tool_count <- as.data.frame(rbind(table(myclusters_df$tnv, myclusters_df$cluster_index)[2,],
                          table(myclusters_df$DVF, myclusters_df$cluster_index)[2,],
                          table(myclusters_df$tv, myclusters_df$cluster_index)[2,],
                          table(myclusters_df$VB, myclusters_df$cluster_index)[2,],
                          table(myclusters_df$VS, myclusters_df$cluster_index)[2,],
                          table(myclusters_df$VS2, myclusters_df$cluster_index)[2,])
                    )

tool_count <- data.frame(t(apply(tool_count, c(1), function(x) {x <- x/table(myclusters_df$cluster_index)})))

tool_count$method <- c("tnv", "DVF", "tv", "VB", "VS", "VS2")

colnames(tool_count) <- c("tool", "cluster_index", "tool_count_norm")
```

ggplot(tool_count, aes(x=tool, y=tool_count_norm,
                   fill=tool,
                   color=tool)) +
  geom_bar(stat="identity") +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "none",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    #legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1)) +
  xlab("Tool") +
  ylab("Proportion of Times in Cluster") + 
  facet_wrap(~cluster_index, nrow=1)
```

```{r}
accuracy_scores_melt <- accuracy_scores %>% 
  select(precision, recall, MCC, numtools, toolcombo) %>%
  group_by(numtools, toolcombo) %>%
  summarise(precision=mean(precision),
            recall=mean(recall),
            MCC=mean(MCC)) %>%
  pivot_longer(cols=c(precision, recall, MCC), 
               names_to="performance_metric",
               values_to="performance_metric_score")
```

```{r}
myclusters_df <- inner_join(accuracy_scores_melt, myclusters_df, 
                            by=c("toolcombo"="combo"))

myclusters_df$cluster_index <- as.factor(myclusters_df$cluster_index)
```

```{r}
ggplot(myclusters_df, aes(x=cluster_index, y=performance_metric_score, 
                                  color=cluster_index, fill=cluster_index)) +
  geom_boxplot() +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  ylab("MCC") +
  xlab("Cluster") +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(9)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(9)), 1)) +
  facet_wrap(~performance_metric)
```
